---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(tidyverse)
```

# Metric Trails

<!-- badges: start -->

<!-- badges: end -->

The goal of Metric Trail is to provide an auditable and reproducible framework for modeling trail running activities from GNSS data such as generated by wearable devices

process GPS generated data such as used in sports apps and wearable devices. You can find functions to load the data and process it. The package can read streams of GPS readings, consisting in latitude, longitude, elevation and time and compute several features for each measure.

GPS data streams can be obtained several ways. The most common file format is GPX files, an XML variant for GPS readings, but fit files (from Garmin devices) and KML files are also supported. Some sports apps like strava offers API services from which streams can be obtained.

For more information about gathering GPX/fit files from strava follow the steps described by [marcusvolz/strava](https://github.com/marcusvolz/strava)

For more information about gathering streams from strava API follow the steps described by [fawda123/rStrava](https://github.com/fawda123/rStrava)

This package contains raw files and sample data as it comes from the latter alternative. Both activity metadata and streams are available, corresponding to activities recorded by the author

### Overview and installation

This is the development repository for GPS_stream, an R package to analyze data from a GPS stream. The development version from [GitHub](https://github.com/) can be installed and loaded as follows:

``` r
# install.packages("remotes")
remotes::install_github("raimun2/metrictrails")
```

## Example

This is a basic example which shows you how to load sample data:

```{r sampledata}
library(GPStream)

data(strava_streams)

act_streams <- strava_streams

```

Data can also be loaded from stream files such as GPX or fit files

```{r read_stream}

gpx_stream <- read_stream("inst/extdata/gpx_activity.gpx")

str(gpx_stream)

fit_stream <- read_stream("inst/extdata/fit_activity.fit")

str(fit_stream)

```

As can be seen, not all files contains the same naming convention for coordinates, so we follow the lat, lon, ele, timestamp, time convention for latitude, longitude, elevation and time signature respectively. This convention is implemented in rename_stream() function.

```{r uniform}
clean_stream1 <- act_streams %>% uniform_stream() %>% filter(id == unique(act_streams$id)[1])

clean_stream2 <- gpx_stream %>% uniform_stream()

clean_stream3 <- fit_stream %>% uniform_stream()

str(clean_stream2)

```

In several occasions elevation readings can be corrupted or missing, so the ele_correction() function is implemented, obtaining elevation data from a digital elevation model (DEM) available in the elevatr package, or in a local DEM if available.

```{r elevation}

correct_elevation <- clean_stream2 %>% ele_correction(replace = FALSE, z = 14)

ggplot(correct_elevation, aes(x=time)) + 
  geom_line(aes(y=ele)) + 
  geom_line(aes(y=ele_DEM), col="red")

```

In other cases, even lat and lon values can be corrupted and therefore need smoothing in order to represent distances correctly. Smoothing can be also applied to interpolate a stream and obtaining an even spaced model of the activity

```{r smooth}

smooth_act <- correct_elevation %>% smooth_stream(interpolate = TRUE, alpha = 0.05, replace = TRUE)

ggplot(correct_elevation, aes(lon, lat)) + 
  geom_point() + 
  geom_point(data = smooth_act, col="red", size=0.7)

```

So far we have corrected coordinates, but we can also estimate different differential features, such as slope, distance, elevation gain/loss, speed, etc. For this purposes is the function differential_stream

```{r differential}

diff_stream <- smooth_act %>% differential_stream()

str(diff_stream)

```

We can also compare two streams to check if they correspond to the same route. The match stream function checks the spatial and sequential continuity of a stream with respect to a reference stream.

For example, with matching streams the function returns the first row and length of the first stream where it matches the second one.

```{r match positive}

data <- clean_stream1
route <- clean_stream2
match_stream(data, route)

```

```{r, echo = FALSE}

ggplot(data, aes(lon, lat)) + 
  geom_point() + 
  geom_point(data = route, col="red", size = 0.7)


```

If we invert the order of the second stream the function detects it

```{r match invert}

data <- clean_stream1
route <- clean_stream2[nrow(clean_stream2):1,]
match_stream(data, route)


```

```{r, echo = FALSE}

ggplot(data, aes(lon, lat)) + 
  geom_point() + 
  geom_point(data = route, col="red", size = 0.7)


```

Also detects if first stream matches partially with second stream

```{r match partial}


data <- clean_stream1
route <- act_streams %>% uniform_stream() %>% 
               filter(id == unique(act_streams$id)[4])
match_stream(data, route)


```

```{r, echo = FALSE}

ggplot(data, aes(lon, lat)) + 
  geom_point() + 
  geom_point(data = route, col="red", size = 0.7)


```

which is different from intersecting streams which diverges on other ends

```{r match overlap}

data <- clean_stream1
route <- act_streams %>% uniform_stream() %>% 
               filter(id == unique(act_streams$id)[15])
match_stream(data, route)
```

```{r, echo = FALSE}

ggplot(data, aes(lon, lat)) + 
  geom_point() + 
  geom_point(data = route, col="red", size = 0.7)


```

Finally it also identifies when streams don't intersect at all

```{r match negative}
data <- clean_stream1
route <- act_streams %>% uniform_stream() %>% 
               filter(id == unique(act_streams$id)[20])
match_stream(data, route)
```

```{r, echo = FALSE}

ggplot(data, aes(lon, lat)) + 
  geom_point() + 
  geom_point(data = route, col="red", size = 0.7)


```

One can also aggregate segments of streams according to distance or time, and using a direct splitting or rolling window. The following code will produce 60 second segments with direct segmentation

```{r time agg}
stream_segmentation <- clean_stream2 %>% differential_stream()

straigth_segments <- agg_stream(stream_segmentation, value = "time", size = 60, windowed = FALSE)

glimpse(straigth_segments)

```

This will produce 100 meters segments with direct segmentation

```{r dist agg}

straigth_segments_d <- agg_stream(stream_segmentation, value = "distance", size = 100, windowed = FALSE)

glimpse(straigth_segments_d)
```

And the following will produce same length segments using rolling window method, see the number of rows produces

```{r window agg}
windowed_segments_d <- agg_stream(stream_segmentation, value = "distance", size = 100, windowed = TRUE)

glimpse(windowed_segments_d)
```

In summary, several functions can be combined to read and analyze GPS streams obtained from physical activities or other sources

```{r closure, message=FALSE, warning=FALSE}

clean_stream <- 
  read_stream("inst/extdata/fit_activity.fit") %>% 
  uniform_stream() %>% 
  ele_correction(replace = FALSE) %>% 
  smooth_stream(interpolate = FALSE, alpha = 0.05, replace = FALSE) %>% 
  differential_stream()
  
ggplot(clean_stream, aes(lon,lat)) +
  geom_point(aes(size=abs(slope), col=hz_velocity)) +
  scale_color_viridis_c()


```